---
title: "Analysis of Time Series-An Introduction"
author: "Rick Dean"
format:
  html:
    toc: false
    toc-depth: 4
    toc-location: "right"
    number-sections: false
    number-offset: 0
    self-contained: true
    smooth-scroll: true
    code-fold: true
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    code-overflow: wrap
    tbl-cap-location: "bottom"
    fig-width: 5
    fig-height: 5
    fig-align: "center"
    fig-cap-location: "bottom"
    minimal: false
    css: ../../style.css
    link-external-newwindow: true
    abstract-title: "Abstract"
    abstract: "The following notes/scripts/plots are inspired by the book [The Analysis of Time Series: An Introduction](http://www.ams.sunysb.edu/~xing/tsRbook/index.html) by Chris Chatfield and Haipeng Xing"  
---

```{r}
#| message: false
#| warning: false

library(data.table)
library(here)
library(magrittr)
library(ggplot2)
library(grid)
library(gtable)
library(tseries)
library(RplotterPkg)
library(RtsaPkg)
```


# 2 Basic Descriptive Techniques

## 2.5 Series that contain a trend and no seasonal variation

### 2.5.2 Filtering
:::topic
Obtain the Beveridge wheat price data set and plot it. Note that the data set `bev` is a time series object of class "ts".
:::

```{r}
#| tbl-cap: Beveridge wheat price data set time series

data(bev, package = "tseries") # set the data
RplotterPkg::create_table(
  x = RtsaPkg::get_series_info(series = bev)$stats,
  container_width_px = 600
)
```

```{r}
#| fig-width: 10
#| fig-cap: Beveridge Wheat Prices

wheat_index_dt <- RtsaPkg::tsObj_to_dt(series = bev)
RplotterPkg::create_scatter_plot(
  df = wheat_index_dt,
  aes_x = "time",
  aes_y = "value",
  x_title = "Year",
  y_title = "Index",
  connect = TRUE
)
```

:::topic
Define a simple moving average plot.
:::

```{r}
#| fig-width: 10
#| fig-cap: Beveridge Wheat Prices, simple moving average 

RtsaPkg::graph_ma(
  df = wheat_index_dt, 
  time_col = "time",
  value_col = "value", 
  window_n = 7,
  x_title = "Year",
  y_title = "Index",
  show_pts = FALSE
)
```

:::topic
Define a weighted moving average plot.
:::

```{r}
#| fig-width: 10 
#| fig-cap: Beveridge Wheat Prices, Weighted Moving Average

RtsaPkg::graph_ma(
  df = wheat_index_dt, 
  time_col = "time",
  value_col = "value",
  ma_type = "wma",
  window_n = 7,
  x_title = "Year",
  y_title = "Index",
  show_pts = FALSE,
)
```

## 2.6 Series that contain a trend and seasonal variation

:::topic
Decompose the quarterly domestic sales of Australian wine time series into seasonal, trend, and irregular components.
:::

```{r}
apath <- file.path(here::here(), "Analysis of Time Series-An Introduction", "data", "aus_wine_sales.csv")
wine_sales_dt <- data.table::fread(apath) %>% 
  data.table::setnames(old = c("V1","V2"), new = c("Date","Sales")) %>% 
  .[, Date := paste(substring(Date, 5, 8), substring(Date, 1, 3), "01", sep = "-")] %>% 
  .[, Date := as.Date(Date, format = "%Y-%b-%d")]
```

```{r}
#| fig-width: 8
#| fig-height: 6
#| warning: false
#| fig-cap: Decomposed Time Series of Domestic Sales of Australian Wine, June 1985 to June 2014 by quarter

wine_sales_ts <- ts(data = wine_sales_dt$Sales, frequency = 4, start = c(1985, 2))
RtsaPkg::graph_decompose(
  series_ts = wine_sales_ts,
  x_title = "Year",
  x_major_date_breaks = "2 year",
  x_date_labels = "%Y"
)
```

## 2.7 Autocorrelation and the Correlogram
### 2.7.2 Interpreting the correlogram

:::topic
Create a random series and plot both the series and its correlogram.
:::  

Create a series data frame from a random normal distribution and plot the autocorrelation (ACF) and partial autocorrelation (PACF):
```{r}
set.seed(1)
random_df <- data.frame (
  time = seq(1:400),
  value = rnorm(400)
)
```

```{r}
#| fig-width: 10 
#| fig-cap: Random Series and Autocorrelations

RtsaPkg::graph_acf(
  df = random_df,
  time_col = "time",
  value_col = "value",
  confid_level = 1.96
)
```

The acf's hover around zero and don't pass the 95% confidence level.

:::topic
Plot the autocorrelation from a series that has a trend and is non-stationary.
:::

Create the non-stationary series.
```{r}
set.seed(1)
# set up the plot dataframe
nonstat_df <- data.frame(
  time = seq(1:400),
  value = cumsum(rnorm(400))
)
```

```{r}
#| fig-width: 10 
#| fig-cap: Non-stationary Random Time Series

RtsaPkg::graph_acf(
  df = nonstat_df,
  time_col = "time",
  value_col = "value",
  confid_level = 1.96
)
```

Appears that the lag at 1 is significant.

:::topic
Plot the autocorrelation of monthly observations on air temperaature in Anchorage, Alaska.
:::

```{r}
apath <- file.path(here::here(), "Analysis of Time Series-An Introduction", "data", "alaska_temp_1951-2016.csv")
alaska_temps_dt <- data.table::fread(apath) %>% 
  .[, Date := paste(substring(as.character(Date), 1, 4), substring(as.character(Date), 5, 6), "01", sep = "-")] %>% 
  .[, Date := as.Date(Date)]
```

```{r}
#| fig-width: 10
#| fig-cap: Autocorrelation for Monthly Alaska Air Tempertures degC

RtsaPkg::graph_acf(
  df = alaska_temps_dt,
  time_col = "Date",
  value_col = "Celsius",
  obs_x_major_date_breaks = "10 year",
  obs_x_date_labels = "%Y",
  confid_level = 1.96
)
```
